Wait-free Queues with Polygarithmic Step Complexity

### 2022-03-24 article/psuedocode.tex
fix garbage collection mistakes

\Statex $\diamondsuit$ \texttt{\textsl{Shared}}
\begin{itemize}
\item \texttt{\textsl{Tree}} to complete
\end{itemize}

@@

\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{enq-left}, sum\textsubscript{enq-left}}
  \textsf{:~~\#enqueues from subblocks in left child, prefix sum of \texttt{num\textsubscript{enq-left}}}
  \item \texttt{\textsl{int} num\textsubscript{deq-left}, sum\textsubscript{deq-left}}
  \textsf{:~~\#dequeues from subblocks in left child, prefix sum of \texttt{num\textsubscript{deq-left}}}
  \item \texttt{\textsl{int} num\textsubscript{enq-right}, \texttt{sum\textsubscript{enq-right}}}
  \textsf{: ~\#enqueues from subblocks in right child, prefix sum of \texttt{num\textsubscript{enq-right}}}
  \item \texttt{\textsl{int} num\textsubscript{deq-right}, \texttt{sum\textsubscript{deq-right}}}
  \textsf{: ~\#dequeues from subblocks in right child, prefix sum of \texttt{num\textsubscript{deq-right}}}
  \item \texttt{\textsl{int} num\textsubscript{enq}, num\textsubscript{deq}}
  \textsf{: \# enqueue, dequeue operations in the block}
  \item \texttt{\textsl{int} sum\textsubscript{enq}, sum\textsubscript{deq}}
  \textsf{: sum of \# enqueue, dequeue operations in blocks up to this one}
  \item \texttt{\textsl{int} num, sum}
  \textsf{: total \# operations in block, prefix sum of \texttt{num}}
  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}
  \textsf{:~~index of the last subblock in the left and right child}
  \item \texttt{\textsl{int} group}
  \textsf{: id of the group of blocks including this propagated together, more precisely the value of the node's counter when propagating this block.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \texttt{\textsl{Root Block} extends \textsl{Block}}
\begin{itemize}
  \item \texttt{\textsl{int} size}
  \textsf{size of queue after this block's operations finish}
  \item \texttt{\textsl{int} sum\textsubscript{non-null deq}}
  \textsf{count of non-null dequeus up to this block}
\end{itemize}

@@

\Function{Object}{Dequeue()}{}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.element= null}
\State \texttt{b.sum\textsubscript{deq}=1}
\State \Call{Append}{b}
\State \texttt{<i, b>=} \Call{Index}{l\textsubscript{pid}, op.loc, 1}
\State \texttt{res=} \Call{ComputeHead}{i, b} \Comment{Index of the enqueue whose argument should be returned}
\State \Return \Call{Get}{res}
\EndFunction{Dequeue}

@@

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{h= n.head}
%\If{\texttt{n.blocks[h]!=null?}} \texttt{h+=1}\EndIf
\State \texttt{c= n.counter}
\State \texttt{<new, c\textsubscript{left}, c\textsubscript{right}>= \Call{CreateBlock}{n, h}}
\State \texttt{new.group= c}
\If{\texttt{new.num==0}} \Return{\texttt{true}}
\ElsIf{\Call{CAS}{n.blocks[h], null, new}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, h+1}}
\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
\EndFor
\State \texttt{\Call{CAS}{n.head, h, h+1}}
\State \Return{\texttt{true}}
\Else
\State \texttt{\Call{CAS}{n.head, h, h+1}}
\State \Return{ \texttt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Function{element}{Get}{\textsl{int} i}
\Comment{Returns $i$th Enqueue.}
\If{\texttt{i }\textbf{is }\texttt{null}}
\State \Return \texttt{null}
\EndIf
\State \texttt{res= \Call{BSearch}{root, sum\textsubscript{enq}, i, 0, root.head}}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\textsubscript{enq}}}
\EndFunction{Get}

@@

\Function{<Block, int, int>}{CreateBlock}{\textsl{node} n, \textsl{int} i} 
\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}. Returns the created block as well as values read from each child counter feild.}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{lastIndex= n.dir.head} \label{lastLine}
\State \texttt{prevIndex= n.blocks[i-1].end\textsubscript{dir}} \label{prevLine}
\State \texttt{lastBlock= n.dir.blocks[lastIndex]}
\State \texttt{prevBlock= n.dir.blocks[prevIndex]}
\State \texttt{c\textsubscript{dir}= n.dir.counter}
\State \texttt{b.end\textsubscript{dir}= lastIndex}
\State \texttt{b.num\textsubscript{enq-dir}= lastBlock.sum\textsubscript{enq} - prevBlock.sum\textsubscript{enq}}
\State \texttt{b.num\textsubscript{deq-dir}= lastBlock.sum\textsubscript{deq} - prevBlock.sum\textsubscript{deq}}
\State \texttt{b.sum\textsubscript{enq-dir}= n.blocks[i-1].sum\textsubscript{enq-dir} + b.num\textsubscript{enq-dir}}
\State \texttt{b.sum\textsubscript{deq-dir}= n.blocks[i-1].sum\textsubscript{deq-dir} + b.num\textsubscript{deq-dir}}
\EndFor
\State \texttt{b.num\textsubscript{enq}= b.num\textsubscript{enq-left} + b.num\textsubscript{enq-right}}
\State \texttt{b.num\textsubscript{deq}= b.num\textsubscript{deq-left} + b.num\textsubscript{deq-right}}
\State \texttt{b.num= b.num\textsubscript{enq} + b.num\textsubscript{deq}}
\State \texttt{b.sum= n.blocks[i-1].sum + b.num}

@@

\begin{algorithm}
\caption{Queue Continued}
\begin{algorithmic}[1]
\setcounter{ALG@line}{83}

\Statex $\leadsto$ \textsf{Precondition:~\texttt{n.blocks[b]} contains $\geq$\texttt{i} enqueues.}
\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
\Comment{\textmd{Returns the \texttt{i}th Enqueue in \texttt{b}th block of node \texttt{n}}}
\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[b]}
\Else
\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{enq-left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
\State \texttt{subBlock= \Call{BSearch}{n.left, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum\textsubscript{enq}} 
\Else
\State \texttt{i= i-n.blocks[b].num\textsubscript{enq-left}}
\State\texttt{subBlock=\Call{BSearch}{n.right, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{right}+1, n.blocks[b].end\textsubscript{right}}}
\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum\textsubscript{enq}} 
\EndIf
\EndIf
\EndFunction{Get}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \texttt{b}th block of node \texttt{n} has propagated up to the root and \texttt{i}th dequeue resides in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{root.blocks[b-1]+i, b}
\Else
\State \texttt{dir= (n.parent.left==n)? left: right}
\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}}
\If{\texttt{dir {\keywordfont is} left}}
\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq-right}}
\Else \State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq} + n.blocks[superBlock].sum\textsubscript{deq-left}}
\EndIf
\State \Return\Call{Index}{n.parent, superBlock, i}
\EndIf
\EndFunction{Index}

@@

\begin{algorithm}
%\caption{Modified Queue \label{algQx}}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Statex $\blacktriangleright$ \texttt{\textsl{PRBTree[RBTNode]} rootBlocks}
  \Statex \textsf{A persistant redblack tree supporting \texttt{append(n),get(i),split(j)}}. \texttt{append(n)} returns \texttt{true} in case successful.

\Statex $\blacktriangleright$ \texttt{\textsl{RBTNode}}
\begin{itemize}

  \item \texttt{\textsl{RootBlock} block}

  \item \texttt{\textsl{int} age}

  \textsf{number of finished operations in the block}

  \item \texttt{\textsl{int} step}

  \textsf{If this node is the root of the RBTree, then step shows the number of appended nodes to the RBTree}

\Function{boolean}{Refresh}{\textsl{node} n} \Comment if n is root
\State \texttt{new=CreateBlock()} \Comment TODO
\If{\texttt{new.num==0}} \Return{\texttt{true}}
\ElsIf{\Call{RBTAppend}{new}}
\Statex..\EndIf
\EndFunction{Refresh}

\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{search in the current RBTRee}
\Else
..
\EndIf
\EndFunction{Index}

\Function{Object}{Dequeue()}{}
\State ..
\State \texttt{N\textsubscript{i}= RBTnode n that n.block contains the invocation of Dequeue}
\State \texttt{N\textsubscript{r}= RBTnode n that n.block contains the response of Dequeue}
\State \texttt{N\textsubscript{i}.age= N\textsubscript{i}.age+1} \Comment{this is a shared counter}
\State \texttt{N\textsubscript{r}.age= N\textsubscript{r}.age+1}
\If{\texttt{N\textsubscript{i} or N\textsubscript{r} become old}} \texttt{update maxOld}
\EndIf
\EndFunction{Dequeue}

\Function{void}{RBTAppend}{block b} \Comment{\textsf{adds block b to the \texttt{rootBlocks}}}
\State \texttt{root= rootBlocks.root}
\If{\texttt{root.step\%$p^2$==0}}
\State \texttt{Help()}
\State \texttt{CollectGarbage()}
\EndIf
\State \texttt{new= RBTnode(b,0,null)}
\State \Return \texttt{rootBlocks.append(new)}
\EndFunction{RBTAppend}
\Statex

\Function{void}{Help}{}\Comment{Helps pending operations}
\State{last= l.head-1}
\For{\texttt{leaf l in leaves}}
\If{\texttt{l.blocks[last] is not null}}
\If{op \textbf{is} DEQ}
\State\texttt{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate()}
\State \texttt{write the response to l.responses[last]}
\EndIf
\EndIf
\EndFor
\EndFunction{Help}
\Statex

>>>>

\Statex $\diamondsuit$ \texttt{\textsl{Shared}}
\begin{itemize}
\item \texttt{\textsl{Tree}} \textit{to complete, how?}
\end{itemize}

@@

\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{enq-left}, sum\textsubscript{enq-left}}
  \textsf{:~~\#enqueues from subblocks in left child, prefix sum of \texttt{num\textsubscript{enq-left}}}
  \item \texttt{\textsl{int} num\textsubscript{deq-left}, sum\textsubscript{deq-left}}
  \textsf{:~~\#dequeues from subblocks in left child, prefix sum of \texttt{num\textsubscript{deq-left}}}
  \item \texttt{\textsl{int} num\textsubscript{enq-right}, \texttt{sum\textsubscript{enq-right}}}
  \textsf{: ~\#enqueues from subblocks in right child, prefix sum of \texttt{num\textsubscript{enq-right}}}
  \item \texttt{\textsl{int} num\textsubscript{deq-right}, \texttt{sum\textsubscript{deq-right}}}
  \textsf{: ~\#dequeues from subblocks in right child, prefix sum of \texttt{num\textsubscript{deq-right}}}
  \item \texttt{\textsl{int} num\textsubscript{enq}, num\textsubscript{deq}}
  \textsf{: \# enqueue, dequeue operations in the block}
  \item \texttt{\textsl{int} sum\textsubscript{enq}, sum\textsubscript{deq}}
  \textsf{: sum of \# enqueue, dequeue operations in blocks up to this one}
  \item \texttt{\textsl{int} num, sum}
  \textsf{: total \# operations in block, prefix sum of \texttt{num}}
  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}
  \textsf{:~~index of the last subblock in the left and right child}
  \item \texttt{\textsl{int} group}
  \textsf{: id of the group of blocks including this propagated together, more precisely the value of the node's counter when propagating this block.}
  \item \texttt{\textsl{int} order}
  \textsf{: the index of the block in the node containing it}
\end{itemize}

@@

\Statex $\blacktriangleright$ \texttt{\textsl{Root Block} extends \textsl{Block}}
\begin{itemize}
  \item \texttt{\textsl{int} size}
  \textsf{: size of queue after this block's operations finish}
  \item \texttt{\textsl{int} sum\textsubscript{non-null deq}}
  \textsf{: count of non-null dequeus up to this block}
  \item \texttt{\textsl{int} age}
  \textsf{: number of finished operations in the block}
\end{itemize}

@@

\Function{Object}{Dequeue()}{}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.element= null}
\State \texttt{b.sum\textsubscript{deq}=1}
\State \Call{Append}{b}
\State \texttt{<i, b>=} \Call{Index}{l\textsubscript{pid}, b.order, 1}
\State \texttt{res=} \Call{ComputeHead}{i, b} \Comment{Index of the enqueue whose argument should be returned} \label{deqRest}
\State \Return \Call{Get}{res}
\State \texttt{b\textsubscript{i}= b}\Comment{block in the root contains the invocation of dequeue}
\State \texttt{b\textsubscript{r}= root.blocks.get(sum\textsubscript{enq}==i)}\Comment{block in the root contains the invocation of dequeue}
\State \texttt{b\textsubscript{i}.age= b\textsubscript{i}.age+1}
\State \texttt{b\textsubscript{r}.age= b\textsubscript{r}.age+1}
\If{\texttt{b\textsubscript{i} or b\textsubscript{r} become old}} \texttt{update maxOld}
\EndIf
\EndFunction{Dequeue}

@@

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{h= n.head}
%\If{\texttt{n.blocks[h]!=null?}} \texttt{h+=1}\EndIf
\State \texttt{c= n.counter}
\State \texttt{<new, c\textsubscript{left}, c\textsubscript{right}>= \Call{CreateBlock}{n, h}}
\State \texttt{new.group= c}
\If{\texttt{new.num==0}} \Return{\texttt{true}}
\ElsIf{\texttt{n} \textbf{is} \texttt{root}}
\If{\texttt{root.blocks.append(new)}}\State\texttt{goto \ref{okcas}}\EndIf
\ElsIf{\Call{CAS}{n.blocks[h], null, new}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}} \label{okcas}
\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, h+1}}
\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
\EndFor
\State \texttt{\Call{CAS}{n.head, h, h+1}}
\State \Return{\texttt{true}}
\Else
\State \texttt{\Call{CAS}{n.head, h, h+1}}
\State \Return{ \texttt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Function{element}{Get}{\textsl{int} i}
\Comment{Returns $i$th Enqueue.}
\If{\texttt{i }\textbf{is }\texttt{null}}
\State \Return \texttt{null}
\EndIf
\State \texttt{res= root.blocks.get(sum\textsubscript{enq}==i).order}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\textsubscript{enq}}}
\EndFunction{Get}

@@

\Function{<Block, int, int>}{CreateBlock}{\textsl{node} n, \textsl{int} i} 
\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}. Returns the created block as well as values read from each child counter feild.}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.order= i}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{lastIndex= n.dir.head} \label{lastLine}
\State \texttt{prevIndex= n.blocks[i-1].end\textsubscript{dir}} \label{prevLine}
\State \texttt{lastBlock= n.dir.blocks[lastIndex]}
\State \texttt{prevBlock= n.dir.blocks[prevIndex]}
\State \texttt{c\textsubscript{dir}= n.dir.counter}
\State \texttt{b.end\textsubscript{dir}= lastIndex}
\State \texttt{b.num\textsubscript{enq-dir}= lastBlock.sum\textsubscript{enq} - prevBlock.sum\textsubscript{enq}}
\State \texttt{b.num\textsubscript{deq-dir}= lastBlock.sum\textsubscript{deq} - prevBlock.sum\textsubscript{deq}}
\State \texttt{b.sum\textsubscript{enq-dir}= n.blocks[i-1].sum\textsubscript{enq-dir} + b.num\textsubscript{enq-dir}}
\State \texttt{b.sum\textsubscript{deq-dir}= n.blocks[i-1].sum\textsubscript{deq-dir} + b.num\textsubscript{deq-dir}}
\EndFor
\State \texttt{b.num\textsubscript{enq}= b.num\textsubscript{enq-left} + b.num\textsubscript{enq-right}}
\State \texttt{b.num\textsubscript{deq}= b.num\textsubscript{deq-left} + b.num\textsubscript{deq-right}}
\State \texttt{b.num= b.num\textsubscript{enq} + b.num\textsubscript{deq}}
\State \texttt{b.sum= n.blocks[i-1].sum + b.num}

@@

\begin{algorithm}
\begin{algorithmic}[1]
\setcounter{ALG@line}{83}

\Statex $\leadsto$ \textsf{Precondition:~\texttt{n.blocks[b]} contains $\geq$\texttt{i} enqueues.}
\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
\Comment{\textmd{Returns the \texttt{i}th Enqueue in \texttt{b}th block of node \texttt{n}}}
\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[b]}
\Else
\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{enq-left}}} \Comment{\texttt{i} exists in the left child of \texttt{n}}
\State \texttt{subBlock= \Call{BSearch}{n.left, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum\textsubscript{enq}} 
\Else
\State \texttt{i= i-n.blocks[b].num\textsubscript{enq-left}}
\State\texttt{subBlock=\Call{BSearch}{n.right, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{right}+1, n.blocks[b].end\textsubscript{right}}}
\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum\textsubscript{enq}} 
\EndIf
\EndIf
\EndFunction{Get}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \texttt{b}th block of node \texttt{n} has propagated up to the root and \texttt{i}th dequeue resides in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{root.blocks.get(order==b-1).sum\textsubscript{deq}+i, b}
\Else
\State \texttt{dir= (n.parent.left==n)? left: right}
\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}}
\If{\texttt{dir {\keywordfont is} left}}
\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq-right}}
\Else \State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq} + n.blocks[superBlock].sum\textsubscript{deq-left}}
\EndIf
\State \Return\Call{Index}{n.parent, superBlock, i}
\EndIf
\EndFunction{Index}

@@

\begin{algorithm}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Statex $\blacktriangleright$ \texttt{\textsl{PRBTree[rootBlock]}}
  \Statex \textsf{A persistant red-black tree supporting \texttt{append(b, key),get(key=i),split(j)}}. \texttt{append(b, key)} returns \texttt{true} in case successful.

\Function{void}{RBTAppend}{block b} \Comment{\textsf{adds block b to the \texttt{root.blocks}}}
\State \texttt{step= root.head}
\If{\texttt{step\%$p^2$==0}}
\State \texttt{Help()}
\State \texttt{CollectGarbage()}
\EndIf
\State \texttt{b.age= 0}
\State \Return \texttt{root.blocks.append(b, b.order)}
\EndFunction{RBTAppend}
\Statex

\Function{void}{Help}{}\Comment{Helps pending operations}
\For{\texttt{leaf l} \textbf{in leaves}}\Comment{\textit{how to iterate over them?}}
\State{\texttt{last= l.head-1}}
\If{\texttt{l.blocks[last]} \textbf{is not} \texttt{null}}
\If{\texttt{l.blocks[last].element==null}} \Comment{operation is dequeue}
\State \texttt{goto \ref{deqRest} with} \Comment{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate()}
\State \texttt{l.responses[last]= response}
\EndIf
\EndIf
\EndFor
\EndFunction{Help}
\Statex
